Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-11-29 review.tex
Now onto the other reviewer. Losing the will to live...

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Measuring contrast processing in the visual system using SSVEP},
  pdfauthor={Alex R. Wade; Daniel H. Baker},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Measuring contrast processing in the visual system using SSVEP}
\author{Alex R. Wade \and Daniel H. Baker}
\date{}

@@

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for over a century. One of the most productive
experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high SNR, and they allow us to analyze interactions between
stimulus components using a convenient mathematical framework. Here we
describe how SSVEPs have been used to study visual contrast over the
past 70 years (Dawson, 1954). Because our thinking about SSVEPs is
well-described by simple mathematical models, we embed code that
illustrates key steps in the modelling and analysis. This paper can
therefore be used both as a review of the use of SSVEP in measuring
human contrast processing, and as an interactive learning aid.

@@

Neurons in the visual areas of the brain are primarily responsive to
changes in cone photoreceptor activations across time and space. This
property, referred to as `contrast', sets the fundamental limits of our
visual abilities, which remain steady over a remarkably wide range of
environmental light levels. The human response to contrast can be
studied using many different techniques. Early work used psychophysical
methods to measure contrast sensitivity (Campbell \& Green, 1965),
defined as the inverse of the lowest contrast that can be reliably
detected. But neural responses can also be measured more directly using
techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG
method known as the steady state visually evoked potential (SSVEP)
technique has contributed to our understanding of human contrast
processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by
visual stimuli flickering at a constant frequency (Regan, 1966). For
contrast-defined stimuli, such as sine-wave gratings, it is strongest at
the occipital pole, adjacent to the early visual areas that generate the
signal, although careful analysis of individual VEPs reveals multiple
generators throughout visual cortex (Di Russo \emph{et al.}, 2005,
2007). The flickering stimulus entrains neural population responses at
multiples of the stimulus frequency, so continuous EEG data are
typically analysed by taking the Fourier transform, and estimating the
amplitude at these frequencies. Two common variants involve sinusoidal
on-off flicker, where the stimulus alternates between a blank background
and the peak contrast, and sinusoidal counterphase flicker, where the
stimulus alternates in phase (i.e the black regions become white and the
white regions become black). On-off flicker can drive independent
populations of on- and off-cells in the retina once per cycle and can
therefore produce a response at the fundamental flicker frequency, known
as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase
flicker contains two transients per cycle and therefore does not produce
a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on.
Because square-waves are spectrally broad-band, square wave flicker
tends to produce more complex spectral harmonics than sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to
reflect nonlinear processing in the visual system (Regan \& Regan,
1988). Because they can arise from different neuronal computations,
different populations and even different stages of the visual system,
different harmonics can exhibit different input-output functions. For
example, 1F and 2F responses can have different thresholds (Bobak
\emph{et al.}, 1984) and scalp topographies (Regan, 1973) while Kaester
et al (Kaestner \emph{et al.}, 2024) show both different CRF slopes and
thresholds for 1F and 2F components generated by dynamic noise.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well (McKeefry \emph{et al.}, 1996; Di
Russo \emph{et al.}, 2001\emph{a}) (see also (Baseler \& Sutter, 1997)).

@@

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the LGN (Levitt
\emph{et al.}, 2001) and cortex (Hubel \& Wiesel, 1959; Movshon, 1975),
will introduce second harmonic components even when the stimulus itself
is modulated in an on-off fashion (McKeefry \emph{et al.}, 1996).
Analogously, in the spatial domain, so-called `simple cells' are
sensitive to the polarity of a spatial contrast modulation while
`complex cells' respond to the presence of patterned spatial contrast
irrespective of its spatial position (Hubel \& Wiesel, 1962). Both the
amplitude and the phase of the SSVEP response to a single
contrast-reversing sine-wave grating therefore contains information
about nonlinear computations performed across a range of retinal and
cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along entire processing pathway is challenging,
careful parametric variation of the input stimuli does allow us to fit
models of early visual visual processing (e.g. (Zemon \& Gordon, 2006))
and, by incorporating mathematical models of gain control mechanisms
with either instantaneous, or temporally-integrated gain control, it is
possible to model gain control using SSVEP data. For example, (Tsai
\emph{et al.}, 2012) demonstrated that a relatively simple gain control
model gave a good account of the pattern of intermodulation responses
produced by two overlaid patterns flickering at different frequencies.
This was achieved by passing full stimulus waveforms through the
transducer nonlinearity, and calculating the Fourier spectrum of the
model output. Our more recent work on signal combination across eyes and
space similarly demonstrated close correspondence between the
predictions of a computational model and empirical data in humans (Baker
\& Wade, 2017). More detailed modelling of intracortical recordings
(Groen \emph{et al.}, 2022) has revealed details of the timecourse of
gain control effects, specifically that normalization is delayed
slightly relative to the initial visual response.

@@

The approach of estimating visual thresholds by extrapolating SSVEP
responses at higher stimulus levels has revealed much about the
development of visual abilities in infants (Harris \emph{et al.}, 1976;
Atkinson \emph{et al.}, 1979; Braddick \emph{et al.}, 1986). In general,
SSVEP measurements of infant vision have revealed that contrast
sensitivity for both achromatic and chromatic contrast as well as
stereoscopic depth perception develops earlier than had been supposed
previously based on behavioural readouts (Dobson \emph{et al.}, 1978;
Norcia \& Tyler, 1985\emph{b}) with both chromatic and achromatic
contrast detection reaching near-adult levels by around six months.
Spatial acuity as measured by SSVEP reaches adult levels more slowly,
but near-adult levels are recorded around one year (Norcia \& Tyler,
1985\emph{b}; Hamer \emph{et al.}, 1989) compared to around six to seven
years with behavioural measures (Atkinson \& Braddick, 1983; Ellemberg
\emph{et al.}, 1999). At least some of this difference is likely due to
the relative objectivity and high SNR of the SSVEP technique compared to
other methods such as preferential looking, which require careful
measurement of the infant's gaze direction. However it should be noted
that other groups have reported electrophysiological correlates of
visual acuity that more closely match the behavioural measures (De
Vries-Khoe \& Spekreijse, 1982).

@@

Amblyopia is a disorder of binocular vision, characterised by one eye
contributing much less to perception than the other. This is often due
to strabismus (squint) or anisometropia (difference in optical
prescription between the eyes) during development. Contemporary accounts
suggest that the amblyopic eye is suppressed by signals from the fellow
eye. SSVEPs provide a convenient and objective method to characterise
the difference in neural response to signals in each eye, and typically
show reduced responses to stimuli in the amblyopic eye (see
Figure~\ref{fig-plotclinical}b) across the contrast range (Baker
\emph{et al.}, 2015; Lygo \emph{et al.}, 2021). There are currently many
novel binocular treatments for amblyopia under development, often
involving virtual reality or stereo display systems designed to
encourage the two eyes to work together. The steady-state approach may
be more sensitive and objective than typical acuity measurements, and
also has the potential to measure suppression between the eyes directly
(e.g. Zheng \emph{et al.}, 2019; Hu \emph{et al.}, 2023; Du \emph{et
al.}, 2023).

Autism is a condition often associated with differences in vision
(Simmons \emph{et al.}, 2009) and other senses (MacLennan \emph{et al.},
2022). Pei \emph{et al.} (2014) used a sweep-VEP method with
counterphase flickering stimuli, and found weaker responses in autistic
children at spatial frequencies around 8c/deg, compared with age-matched
controls. This was subsequently replicated in a further pediatric sample
by Vilidaite \emph{et al.} (2018) (see Figure~\ref{fig-plotclinical}c),
who additionally found weaker responses in autistic adults at the second
harmonic (using on/off flicker). Interestingly this study replicated its
key findings in a \emph{Drosophila} genetic model of autism (\emph{Nhe3}
mutations), illustrating the translational potential of the steady-state
approach, as well as identifying a possible biomarker for autism.

@@

Attention exerts a profound influence on visual performance: for
example, instructing people to attend covertly to a spatial location
improves their performance on a target detection task significantly
(Bashinski \& Bacharach, 1980; Posner, 1980; Carrasco \emph{et al.},
2000; Cameron \emph{et al.}, 2002; Morrone \emph{et al.}, 2004; Pestilli
\emph{et al.}, 2009). In principle, this enhancement may be driven both
by modulation of the underlying signal or noise characteristics, or by
additional decision-theoretic factors such as reduction in spatial
uncertainty (Petrov \emph{et al.}, 2006; Gould \emph{et al.}, 2007).
Early experiments with non-human primates showed little evidence for
attentionally-driven changes in neuronal spike rates (Luck \emph{et
al.}, 1997; McAdams \& Maunsell, 1999; Mehta \emph{et al.},
2000\emph{a}, 2000\emph{b}; Marcus \& Van Essen, 2002), but with the
advent of spatially-resolved human brain imaging methods in the late
1990s it became apparent that spatial attention was linked to frank
changes in both fMRI BOLD responses (Tootell \emph{et al.}, 1998; Somers
\emph{et al.}, 1999; Gandhi \emph{et al.}, 1999; Brefczynski \& DeYoe,
1999; Kastner \emph{et al.}, 1999; Buracas \& Boynton, 2007; Silver
\emph{et al.}, 2007; Li \emph{et al.}, 2008; Murray, 2008) and EEG
signals (Morgan \emph{et al.}, 1996; Müller \emph{et al.}, 1998; Müller
\& Hillyard, 2000; Ding \emph{et al.}, 2006).

Electrophysiological and psychophysical measurements of the effect of
attention on both luminance and chromatic contrast have strongly
implicated gain control as an underlying mechanism (Lu \& Dosher, 1998;
Di Russo \& Spinelli, 1999\emph{a}; Di Russo \emph{et al.},
2001\emph{b}). These effects also appear to differ between chromatic and
achromatic pathways (Di Russo \& Spinelli, 1999\emph{b}) - perhaps as a
result of the different levels of nonlinear gain control in the early
pre-cortical magno-, parvo- and konio-cellular pathways (Derrington \&
Lennie, 1984; Kaplan \& Shapley, 1986; Lee \emph{et al.}, 1990; Solomon
\& Lennie, 2005). In the late 2000s a comprehensive theoretical model
for attentional modulation was developed that framed it as a gain
control computation (Reynolds \& Heeger, 2009; Boynton, 2009). This
framework has proven to be influential - explaining a wide range of
phenomena from the earlier literature and demonstrating subtle
interactions between the size of the attentional `spotlight' and the
stimulus configuration which rationalise many apparent contradictions in
the literature. Direct measurements of attentional modulation of
achromatic SSVEP signals are broadly consistent with this model
(Lauritzen \emph{et al.}, 2010; Hou \emph{et al.}, 2016; Martinovic \&
Andersen, 2018) and confirm the relatively weaker role of spatial
attention on responses driven by chromatic stimuli - particularly those
that isolate the opponent S-(L+M) cone pathway (Highsmith \& Crognale,
2010; Wang \& Wade, 2011). The gain control model of attention can also
be extended to SSVEP studies of feature-based attention which show that
the modulatory effects can be targeted to the most informative neuronal
populations (Verghese \emph{et al.}, 2012).

@@

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)},
  pdfauthor={Alex R. Wade; Daniel H. Baker},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Measuring contrast processing in the visual system using the
steady state visually evoked potential (SSVEP)}
\author{Alex R. Wade \and Daniel H. Baker}
\date{}

@@

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for over a century. One of the most productive
experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high signal to noise ratio (SNR), and they allow us to
analyze interactions between stimulus components using a convenient
mathematical framework. Here we describe how SSVEPs have been used to
study visual contrast over the past 70 years (Dawson, 1954). Because our
thinking about SSVEPs is well-described by simple mathematical models,
we embed code that illustrates key steps in the modelling and analysis.
This paper can therefore be used both as a review of the use of SSVEP in
measuring human contrast processing, and as an interactive learning aid.

@@

Neurons in the visual areas of the brain are primarily responsive to
changes in cone photoreceptor activations across time and space. This
property, referred to as `contrast', sets the fundamental limits of our
visual abilities, which remain steady over a remarkably wide range of
environmental light levels. The human response to contrast can be
studied using many different techniques. Early work used psychophysical
methods to measure contrast sensitivity Schade (1956), defined as the
inverse of the lowest contrast that can be reliably detected. But neural
responses can also be measured more directly using techniques such as
functional magnetic resonance imaging (fMRI), magnetoencephalography
(MEG), and electroencephalography (EEG). Here we will describe how an
EEG method known as the steady state visually evoked potential (SSVEP)
technique has contributed to our understanding of human contrast
processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by
visual stimuli flickering at a constant frequency (Regan, 1966\emph{a}).
For contrast-defined stimuli, such as sine-wave gratings, it is
strongest at the occipital pole, adjacent to the early visual areas that
generate the signal, although careful analysis of individual
visually-evoked potentials (VEPs) reveals multiple generators throughout
visual cortex (Di Russo \emph{et al.}, 2005, 2007). The flickering
stimulus entrains neural population responses at multiples of the
stimulus frequency, so continuous EEG data are typically analysed by
taking the Fourier transform, and estimating the amplitude at these
frequencies.

Two common stimulus variants involve sinusoidal on-off (or
`appearance/disappearance') flicker, where the stimulus alternates
between a blank background and the peak contrast, and sinusoidal
counterphase flicker, where the stimulus alternates in phase (i.e the
black regions become white and the white regions become black). On-off
flicker can excite independent populations of on- and off-cells in the
visual system once per cycle. For spatial patterns, the contributions of
individual cortical cells to these excitations are thought to sum at the
scalp and generate an average of their onset responses on each cycle.
For very low spatial frequencies (including zero) resposes at the scalp
will be dominated by a single on- or off- cell type on each half cycle.

If the amplitudes and waveforms of responses to the stimulus appearance
and disappearnce are not perfectly balanced, a response at the
fundamental flicker frequency, known as 1F, and its integer harmonics
can be produced. For stimuli of any spatial frequency (including zero),
an imbalance can be caused by differences in either the onset- and
offset- responses of either on- or off-cells. For example, if only
on-sensitive neurons were present (firing to the onset, but not the
offset of local increases in luminance) the response to the on/off (or
`appearance/disappearance') presentation of a grating of any spatial
frequency below the resolution limit would consist entirely of odd
harmonics.

By comparison, counterphase flickering patterns generate two essentially
identical transients per cycle and therefore does not produce a response
at 1F, only even harmonics: 2F, 4F, 6F and so on. Because square-waves
are spectrally broad-band, square wave flicker (either on/off or
contrast reversig) tends to produce additional spectral harmonics
compared to sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to
reflect nonlinear processing in the visual system (Regan \& Regan,
1988). Because they can arise from different neuronal computations,
different populations and even different stages of the visual system,
different harmonics can exhibit different input-output functions. For
example, 1F and 2F responses can have different thresholds (Bobak
\emph{et al.}, 1984) and scalp topographies (Regan, 1973) while Kaester
et al (Kaestner \emph{et al.}, 2024) show both different response slopes
and thresholds for 1F and 2F components generated by dynamic noise.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well McKeefry \emph{et al.} (1996) (see
also (Baseler \& Sutter, 1997)).

@@

Even without considering a spatial component, the early visual system is
far more complex than the model here suggests. For example as well as
cells that code positive or negative contrast in a more or less
continuous manner, the retina also contains `transient' cells that code
temporal changes in contrast. These cells (Kuffler \emph{et al.}, 1957;
Alpern, 1971), and cells with similar properties in the lateral
geniculate nucleus (LGN) (Levitt \emph{et al.}, 2001) and cortex (Hubel
\& Wiesel, 1959; Movshon, 1975), will introduce second harmonic
components even when the stimulus itself is modulated in an on-off
fashion (McKeefry \emph{et al.}, 1996). Analogously, in the spatial
domain, so-called `simple cells' are sensitive to the polarity of a
spatial contrast modulation while `complex cells' respond to the
presence of patterned spatial contrast irrespective of its spatial
position (Hubel \& Wiesel, 1962). Both the amplitude and the phase of
the SSVEP response to a single contrast-reversing sine-wave grating
therefore contains information about nonlinear computations performed
across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along entire processing pathway is challenging, and
a full system identification may require the use of multiple different
input frequencies (Boyd \emph{et al.}, 1983), careful parametric
variation of the input stimuli does allow us to fit models of early
visual visual processing (e.g. (Zemon \& Gordon, 2006)) and, by
incorporating mathematical models of gain control mechanisms with either
instantaneous, or temporally-integrated gain control, it is possible to
model gain control using SSVEP data. For example, (Tsai \emph{et al.},
2012) demonstrated that a relatively simple gain control model gave a
good account of the pattern of intermodulation responses produced by two
overlaid patterns flickering at different frequencies. This was achieved
by passing full stimulus waveforms through the transducer nonlinearity,
and calculating the Fourier spectrum of the model output. More recent
work has used this approach to study binocular integration in both
normal subjects and amblyopesHou \emph{et al.} (2020). Our more recent
work on signal combination across eyes and space similarly demonstrated
close correspondence between the predictions of a computational model
and empirical data in humans (Baker \& Wade, 2017). More detailed
modelling of intracortical recordings (Groen \emph{et al.}, 2022) has
revealed details of the timecourse of gain control effects, specifically
that normalization is delayed slightly relative to the initial visual
response.

@@

The approach of estimating visual thresholds by extrapolating SSVEP
responses at higher stimulus levels has revealed much about the
development of visual abilities in infants (Harris \emph{et al.}, 1976;
Atkinson \emph{et al.}, 1979; Braddick \emph{et al.}, 1986). In general,
SSVEP measurements of infant vision have revealed that contrast
sensitivity and acuity for both achromatic and chromatic stimuli as well
as stereoscopic depth perception develops earlier than had been supposed
previously based on behavioural readouts (Dobson \emph{et al.}, 1978;
Norcia \& Tyler, 1985\emph{b}). For example, using SSVEPs, Norcia and
Tyler (Norcia \& Tyler, 1985\emph{b}) found acuity thresholds in one
month old infants that were roughly half those estimated previously from
behavioural experiments (reviewed in (Brown, 1990)). The ability to make
very sensitive measurements of early visual function in infants also
allows researchers to monitor the trajectory of visual development and
identify different temporal epochs of developmental change (Norcia
\emph{et al.}, 1990).

SSVEP measures of thresholds for detecting purely chromatic stimuli find
responses to isoluminant stimuli by around 5 to six weeks with contrast
detection reaching near-adult levels by around six months (Morrone
\emph{et al.}, 1993). Spatial acuity as measured by SSVEP reaches adult
levels more slowly, but near-adult levels are recorded around one year
(Norcia \& Tyler, 1985\emph{b}; Hamer \emph{et al.}, 1989) compared to
around six to seven years with behavioural measures (Atkinson \&
Braddick, 1983; Ellemberg \emph{et al.}, 1999).

At least some of this difference is likely due to the relative
objectivity and high SNR of the SSVEP technique compared to other
methods such as preferential looking, which require careful measurement
of the infant's gaze direction. In slightly older chilren, other groups
have reported electrophysiological correlates of visual acuity that more
closely match the behavioural measures (De Vries-Khoe \& Spekreijse,
1982).

@@

Amblyopia is a disorder of binocular vision, characterised by one eye
contributing much less to perception than the other. This is often due
to strabismus (squint) or anisometropia (difference in optical
prescription between the eyes) during development. Contemporary accounts
suggest that the amblyopic eye is suppressed by signals from the fellow
eye. SSVEPs provide a convenient and objective method to characterise
the difference in neural response to signals in each eye, and typically
show reduced responses to stimuli in the amblyopic eye (see
Figure~\ref{fig-plotclinical}b) across the contrast range (Baker
\emph{et al.}, 2015; Lygo \emph{et al.}, 2021). Measurments of
interocular contrast modulation (for example, Hou 侯川 \emph{et al.}
(2021) can also provide information about the way that amblyopia changes
the balance of excitation and inhibition between eyes.

There are currently many novel binocular treatments for amblyopia under
development, often involving virtual reality or stereo display systems
designed to encourage the two eyes to work together. The steady-state
approach may be more sensitive and objective than typical acuity
measurements, and also has the potential to measure suppression between
the eyes directly (e.g. Zheng \emph{et al.}, 2019; Hu \emph{et al.},
2023; Du \emph{et al.}, 2023).

Autism is a condition often associated with differences in vision
(Simmons \emph{et al.}, 2009) and other senses (MacLennan \emph{et al.},
2022). Pei \emph{et al.} (2014) used a sweep-VEP method with
counterphase flickering stimuli, and found weaker responses in autistic
children at spatial frequencies around 8c/deg, compared with age-matched
controls. This was subsequently extended to measurements of contrast
sensitivity in a further pediatric sample by Vilidaite \emph{et al.}
(2018) (see Figure~\ref{fig-plotclinical}c), who additionally found
weaker responses in autistic adults at the second harmonic (using on/off
flicker). Interestingly this study replicated its key findings in a
\emph{Drosophila} genetic model of autism (\emph{Nhe3} mutations),
illustrating the translational potential of the steady-state approach,
as well as identifying a possible biomarker for autism.

@@

Attention exerts a profound influence on visual performance: for
example, instructing people to attend covertly to a spatial location
improves their performance on a target detection task significantly
(Bashinski \& Bacharach, 1980; Posner, 1980; Carrasco \emph{et al.},
2000; Cameron \emph{et al.}, 2002; Morrone \emph{et al.}, 2004; Pestilli
\emph{et al.}, 2009). In principle, this enhancement may be driven both
by modulation of the underlying signal or noise characteristics, or by
additional decision-theoretic factors such as reduction in spatial
uncertainty (Petrov \emph{et al.}, 2006; Gould \emph{et al.}, 2007).
Early experiments with non-human primates showed little evidence for
attentionally-driven changes in neuronal spike rates (Luck \emph{et
al.}, 1997; McAdams \& Maunsell, 1999; Mehta \emph{et al.},
2000\emph{a}, 2000\emph{b}; Marcus \& Van Essen, 2002), but with the
advent of spatially-resolved human brain imaging methods in the late
1990s it became apparent that spatial attention was linked to frank
changes in both fMRI (Tootell \emph{et al.}, 1998; Somers \emph{et al.},
1999; Gandhi \emph{et al.}, 1999; Brefczynski \& DeYoe, 1999; Kastner
\emph{et al.}, 1999; Buracas \& Boynton, 2007; Silver \emph{et al.},
2007; Li \emph{et al.}, 2008; Murray, 2008) and EEG signals (Morgan
\emph{et al.}, 1996; Müller \emph{et al.}, 1998; Müller \& Hillyard,
2000; Ding \emph{et al.}, 2006).

Electrophysiological and psychophysical measurements of the effect of
attention on both luminance and chromatic contrast have strongly
implicated gain control as an underlying mechanism (Lu \& Dosher, 1998;
Di Russo \& Spinelli, 1999\emph{a}; Di Russo \emph{et al.},
2001\emph{b}). These effects also appear to differ between chromatic and
achromatic pathways (Di Russo \& Spinelli, 1999\emph{b}) - perhaps as a
result of the different levels of nonlinear gain control in the early
pre-cortical magno-, parvo- and konio-cellular pathways (Derrington \&
Lennie, 1984; Kaplan \& Shapley, 1986; Lee \emph{et al.}, 1990; Solomon
\& Lennie, 2005). In the late 2000s a comprehensive theoretical model
for attentional modulation was developed that framed it as a contrast
gain control computation (Reynolds \& Heeger, 2009; Boynton, 2009). This
framework has proven to be influential - explaining a wide range of
phenomena from the earlier literature and demonstrating subtle
interactions between the size of the attentional `spotlight' and the
stimulus configuration which rationalise many apparent contradictions in
the literature - in particular reports of a response- rather than a
contrast-gain control phenomenon. Direct measurements of attentional
modulation of achromatic SSVEP signals are broadly consistent with this
model (Lauritzen \emph{et al.}, 2010; Hou \emph{et al.}, 2016;
Martinovic \& Andersen, 2018) and confirm the relatively weaker role of
spatial attention on responses driven by chromatic stimuli -
particularly those that isolate the opponent S-(L+M) cone pathway
(Highsmith \& Crognale, 2010; Wang \& Wade, 2011). The gain control
model of attention can also be extended to SSVEP studies of
feature-based attention which show that the modulatory effects can be
targeted to the most informative neuronal populations (Verghese \emph{et
al.}, 2012).

@@

\bibitem[\citeproctext]{ref-norciaChangesCorticalActivity2000}
Norcia AM, Harrad RA \& Brown RJ (2000). Changes in cortical activity
during suppression in stereoblindness. \emph{NeuroReport} \textbf{11,}
1007.
